---
title: "Multiple Linear Regression models"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "README_files/1-MLR/")
```


## Chicago example exercise

Trip production of 57 Traffic Assignment Zones of Chicago in 1960's

> Your task: Estimate a linear regression model that predicts trips per occupied dwelling unit.  

### Variables:

* `TODU`: Motorized Trips (private car or Public Transportation) per occupied dwelling unit;
* `ACO`: Average car ownership (cars per dwelling);
* `AHS`: Average household size;
* `SRI`: Social Rank Index:  
      1. proportion of blue-collar workers (e.g., construction, mining);  
      2. proportion of people with age higher than 25 years that have completed at least 8 year of education;
        (_**Note:** The SRI has its maximum value when there are no blue-collar workers and all adults have education            of at least 8 years_) 
* `UI`: Urbanization Index:  
      1. fertility rate, defined as the ratio of children under 5 years of age to  the female population of childbearing age;  
      2. female labor force participation rate, meaning the % of women who are in the labor force;  
      3. % of single family units to total dwelling units.

    The degree of urbanization index would be increased by
        a) lower fertility rate,
        b) higher female labor force participation rate, and
       c) higher proportion of single dwelling units.
      (_**Note:** High values for this index imply less attachment to the home_)

* `SI`:Segregation Index
   It measures the proportion of an area to which minority groups (e.g: non-whites, foreign-born, Eastern Europeans) live in isolation.
     (_**Note:** High values for this index imply that those communities are less prone to leaving their living areas and as such to having lower levels of mobility_)



#### Import Libraries
Let's begin!

For the first time, you will need to install some of the packages. 
Step by step: 

  1. Go to Packages on the lower right display window and click install
  2. Write the library you want to install and click "install"
  
Or... `install.packages("readxl","tidyverse")` etc...

Depending on the version of your R, `DataExplorer` may need to be installed from source, such as
```r
if (!require(devtools)) install.packages("devtools")
devtools::install_github("boxuancui/DataExplorer")
```

Now, import these libraries:
```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(readxl) #Library used to import excel files
library(tidyverse) # Library used in data science to perform exploratory data analysis
library(skimr) # Library used for providing a summary of the data
library(DataExplorer) # Library used in data science to perform exploratory data analysis
library(corrplot) # Library used for correlation plots
library(car) # Library used for testing autocorrelation (Durbin Watson)
library(olsrr) # Library used for testing multicollinearity (VIF, TOL, etc.)
```

### EXPLORATORY DATA ANALYSIS (EDA)

#### Import dataset

```{r}
dataset <- read_excel("Data/TDM_Class3_MLR_Chicago_Example.xls") 
```

#### Check the structure of the dataset 
```{r}
str(dataset)
```

#### Take a first look at the dataset

```{r paged.print=TRUE}
head(dataset, 10)
```

#### Check the type and class of the dataset

```{r echo=TRUE, eval=FALSE} 
typeof(dataset)
class(dataset)
```

```{r echo=FALSE, purl = FALSE}
typeof(dataset)
class(dataset)
```

#### Transform the dataset into a dataframe

```{r}
df <- data.frame(dataset)
```

#### Compare the structure of the dataset with df

```{r}
str(dataset)
str(df)
```

> **Note:** The dataframe function transforms columns into variables and rows into observations. 

#### Take a look at the dataframe

```{r}
head(df, 10)
```

#### Show summary statistics
 
```{r}
skim(df)
```

#### Check missing data

```{r}
plot_missing(df)
```

#### Plot histograms of all the continuous variables

```{r}
plot_histogram(df)
```

> **Note**: Take a special look at TODU, and see if the variable looks like a normal distribution.  

#### Plot boxplots of each independent variable with TODU

```{r}
plot_boxplot(df, by = "TODU")
```

> **Note**: If you increase the average car ownership (ACO) it will tend to increase the number of trips per dwelling unit (TODU). This makes sense. Try analyzing the other relations and check if it is coherent. 

#### Plot correlation heatmaps

```{r}
res1 <- cor.mtest(df, conf.level = .95)

corrplot(cor(df), p.mat = res1$p, method = "number", type = "upper", order="hclust", sig.level = 0.05)
```


> **Note:** try putting into method "color" or "circle", and see the diference.

> **Note:** The pairwise correlations that are crossed are statistically insignificant.The null hypothesis is that correlation is zero.This means that the correlations are only significant when you reject the null hypothesis (pvalue < 0.05). 

Therefore, take a look at this example and check the pvalue of a crossed pair correlation: 

```{r}
cor.test(df$AHS, df$SI)
```

> **Note:** Correlation heatmaps only consider pairwise correlations and does not demonstrate multicollinearity. 

### MULTIPLE LINEAR REGRESSION

> y(`TODU`) = bo + b1*`ACO` + b2*`AHS` + b3*`SI` + b4*`SRI` +b5*`UI` + Error

#### Before running the model, you need to check if the assumptions are met. For instance, let's take a look if the independent variables have linear relation with the dependent variable.

```{r}
plot(x = df$TODU, y = df$ACO, xlab = "TODU", ylab = "ACO")  
plot(x = df$TODU, y = df$AHS, xlab = "TODU", ylab = "AHS")  
plot(x = df$TODU, y = df$SI, xlab = "TODU", ylab = "SI")  
plot(x = df$TODU, y = df$SRI, xlab = "TODU", ylab = "SRI")  
plot(x = df$TODU, y = df$UI, xlab = "TODU", ylab = "UI")
```

Or you could execute a pairwise scatterplot matrix, that compares every variable with each other: 

```{r}
pairs(df[,1:6], pch = 19, lower.panel = NULL)
```

> **Note:** SRI and TODU do not have a linear relationship. This should interfere on the model. 

#### Check if the Dependent variable is normally distributed

If the sample is smaller than 2000 observations, use Shapiro-Wilk test: 

```{r}
shapiro.test(df$TODU)
```

If not, use the Kolmogorov-Smirnov test

```{r}
ks.test(df$TODU, "pnorm", mean=mean(df$TODU), sd = sd(df$TODU))
```

> **Note:** The warning that appears in the Kolmogorov-Smirnov test:
"ties should not be present for the Kolmogorov-Smirnov test".
Most likely what happened is that this test is only reliable with continuous variables. Although TODU is a continuous variable, the small sample size, makes it likely to have repeated values. Consequently, the test considers TODU as a categorical variable. Therefore, this is another evidence, that for small samples it is more appropriate to use the Shapiro-Wilk Test.

> **Note:** The null hypothesis of both tests is that the distribution is normal. Therefore, for the distribution to be normal, the pvalue > 0.05 and you should not reject the null hypothesis. 

#### Finally, let's run the multiple linear regression model!

```{r}
model <- lm(TODU ~ ACO + AHS + SI + SRI + UI, data = df)
summary(model)
```

> **Note**: First check the pvalue and the F statistics of the model to see if there is any statistical relation between the Dependent Variable and the Independent Variables. If pvalue < 0.05 and the F statistics > Fcritical = 2,39, then the model is statistically acceptable.

> **Note**: The Rsquare and Adjusted Rsquare evaluate the amount of variance that is explained by the model. The diference between one and another is that the Rsquare does not consider the number of variables. If you increase the number of variables in the model, the Rsquare will tend to increase which can lead to overfitting. On the other hand, the Adjusted Rsquare adjust to the number of independent variables. 

> **Note**: Take a look a the tvalue and the Pr(>|t|). If the tvalue > 1,96 or Pr(>|t|) < 0,05, then the IV is statistically significant to the model. 

> **Note**: To analyze the estimates of the variables, you should first check the signal and evaluate if the independent variable has a direct or inverse relationship with the dependent variable. It is only possible to evaluate the magnitude of the estimate if all variables are continuous and standarzized or by calculating the elasticities. The elasticities are explained and demonstrated in chapter 4. 

```{r}
plot(model)
```

* **Residuals vs Fitted:** This plot is used to detect non-linearity, heteroscedasticity, and outliers. 

* **Normal Q-Q:** The quantile-quantile (Q-Q) plot is used to check if the dependent variable follows a normal distribution.

* **Scale-Location:** This plot is used to verify if the residuals are spread equally (homoscedasticity) or not (heteroscedasticity) through the sample. 

* **Residuals vs Leverage:** This plot is used to detect the impact of the outliers in the model. If the outliers are outside the Cook-distance, this may lead to serious problems in the model. 

Try analyzing the plots and check if the model meets the assumptions. 

#### Execute the Durbin Watson test to evaluate autocorrelation of the residuals

```{r}
durbinWatsonTest(model)
```

> **Note:** In the Durbin-Watson test, values of the D-W Statistic vary from 0 to 4. If the values are from 1.8 to 2.2 this means that there is no autocorrelation in the model. 

#### Calculate the VIF and TOL to test multicollinearity

```{r}
ols_vif_tol(model)
```

> **Note:** Values of VIF > 5, indicate multicollinearity problems.

#### Calculate the Condition Index to test multicollinearity

```{r}
ols_eigen_cindex(model)
```

> **Note:** Condition index values > 15 indicate multicollinearity problems, and values > 30 indicate serious problems of multicollinearity.

#### To test both simultaneously, you can run the code below:

```{r}
ols_coll_diag(model)
```


<!-- knitr::purl("1-MultipleLinearRegression.Rmd", "Code/1-MultipleLinearRegression.R", documentation = 2) -->

