---
title: "Exploratory Factor Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "README_files/2-FactorAnalysis/")
```

## EXAMPLE EXERCISE: 
### "Residential location satisfaction in the Lisbon metropolitan area"


The aim of this study was to examine the perception of households towards their residential location considering several land use and accessibility factors as well as household socioeconomic and attitudinal characteristics.

_Reference:_ **Martinez, L. G., de Abreu e Silva, J., & Viegas, J. M. (2010). Assessment of residential location satisfaction in the Lisbon metropolitan area, TRB (No. 10-1161).**

> Your task: Analyse the data and create meaningful latent factors.  

### Variables: 

* `DWELCLAS`: Classification of the dwelling;
* `INCOME`: Income of the household;
* `CHILD13`: Number of children under 13 years old;
* `H18`: Number of household members above 18 years old;
* `HEMPLOY`: Number of household members employed;
* `HSIZE`: Household size;
* `IAGE`: Age of the respondent;
* `ISEX`: Sex of the respondent;
* `NCARS`: Number of cars in the household;
* `AREA`: Area of the dwelling;
* `BEDROOM`: Number of bedrooms in the dwelling;
* `PARK`: Number of parking spaces in the dwelling; 
* `BEDSIZE`: BEDROOM/HSIZE;
* `PARKSIZE`: PARK/NCARS;
* `RAGE10`: 1 if Dwelling age <= 10;
* `TCBD`: Private car distance in time to CBD;
* `DISTTC`: Euclidean distance to heavy public transport system stops;
* `TWCBD`: Private car distance in time of workplace to CBD;
* `TDWWK`: Private car distance in time of dwelling to work place;
* `HEADH`: 1 if Head of the Household;
* `POPDENS`: Population density per hectare;
* `EQUINDEX`: Number of undergraduate students/Population over 20 years old (500m)


### Rules of thumb: 
* At least 10 variables
* n < 50 (Unacceptable); n > 200 (recommended)
* It is recommended to use continuous variables. If your data contains categorical variables, you should transform them to dummy variables. 

### Assumptions: 
* Normality; 
* linearity; 
* Homogeneity;
* Homoscedasticity (some multicollinearity is desirable);
* Correlations between variables < 0.3 (not appropriate to use Factor Analysis)

#### Import Libraries

Let's start!

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
library(foreign) # Library used to read SPSS files
library(nFactors) # Library used for factor analysis
library(tidyverse) # Library used in data science to perform exploratory data analysis
library(summarytools) # Library used for checking the summary of the dataset
library(psych) # Library used for factor analysis
library(GPArotation) # Library used for factor analysis
```


#### Set working directory

```{r}
setwd("G:/O meu disco/TDM - Lecture R/TDM github/Transport-Demand-Modelling/")
```

#### Import dataset

```{r message=FALSE, warning=FALSE}
data <- read.spss("Data/example_fact.sav")
```

#### Take a look at the main characteristics of the dataset

```{r}
str(data)
```

#### Transform dataset into dataframe

```{r}
df <- data.frame(data)
```

#### Check summary statistics of variables

```{r}
descriptive_stats <- dfSummary(df, style='grid', plain.ascii = FALSE, graph.col = TRUE)

view(descriptive_stats)

summary(df)
```

> **Note:** I used other functions for summary statistics in last exercise.

####Take a look at the dataset

```{r}
head(df,10)
summary(df)
```

#### Make ID as row names or case number

```{r}
df<-data.frame(df, row.names = 1)
```

#### Evaluating the assumptions for factoral analysis: 

#### Let's run a random regression model in order to evaluate some assumptions

```{r}
random = rchisq(nrow(df), 32)
fake = lm(random ~ ., data = df)
standardized = rstudent(fake)
fitted = scale(fake$fitted.values)
```

* **Normality**

```{r}
hist(standardized)
```

* **Linearity**

```{r}
qqnorm(standardized)
abline(0,1)
```

* **Homogeneity** 

```{r}
plot(fitted, standardized)
abline(0,0)
abline(v=0)
```

#### Examine correlations
```{r}
corr_matrix <- cor(df, method = "pearson")
corr_matrix
```


#### Check for correlation adequacy (pattern between variables) - Bartlett's Test

```{r}
cortest.bartlett(corr_matrix, n = nrow(df))
```

#### Check for sampling adequacy - KMO test

```{r}
KMO(corr_matrix)
```

> **Note:** We want at least 0.7 of the overall Mean Sample Adequacy (MSA)

#### Determine the number of factors to extract

**1._Parallel Analysis_**

```{r}
num_factors = fa.parallel(df, fm = "ml", fa = "fa")
```

> **Note:** fm = factor math; ml = maximum likelihood; fa = factor analysis

**2. _Kaiser Criterion_** 

```{r}
sum(num_factors$fa.values > 1)
```

> **Note:** Determines the number of factors with eigenvalue > 1

**3. _Principal Component Analysis_ (PCA)**

```{r}
df_pca <- princomp(df, cor=TRUE)
```
    
> **Note:** cor = TRUE, standardizes your dataset

  # print variance that explains the components
```{r}
summary(df_pca)  
loadings(df_pca) 
```

Scree Plot


plot(df_pca,type="lines") 
fit$scores 
#biplot(df_pca)



# EXPLORATORY FACTOR ANALYSIS

## Model 1: No rotation

df_factor <- factanal(df, factors = 4, rotation = "none", scores=c("regression"), fm = "ml")


## Model 2: Rotation Varimax

df_factor_var <- factanal(df, factors = 4, rotation = "varimax", scores=c("regression"), fm = "ml")


## Model 3: Rotation Oblimin

df_factor_obl <- factanal(df, factors = 4, rotation = "oblimin", scores=c("regression"), fm = "ml")



# Let's print out the results of df_factor_obl, and take a look. 

print(df_factor, digits=2, cutoff=0.3, sort=TRUE)

## Note: We used a cutoff of 0.3 due to the sample size is higher than 350 obs.

## Note: Variability contained in the factors = Communality + Uniqueness

## Note: Varimax assigns orthogonal rotation, 
## and oblimin assigns oblique rotation.


#Plot factor 1 against factor 2, and compare the results of different rotations

## No Rotation
plot(df_factor$loadings[,1], 
     df_factor$loadings[,2],
     xlab = "Factor 1", 
     ylab = "Factor 2", 
     ylim = c(-1,1),
     xlim = c(-1,1),
     main = "No rotation")
abline(h = 0, v = 0)
load <- df_factor$loadings[,1:2]
text(text(load,labels=names(df),cex=.7, col="blue"))
abline(h = 0, v = 0)

## Varimax rotation
plot(df_factor_var$loadings[,1], 
     df_factor_var$loadings[,2],
     xlab = "Factor 1", 
     ylab = "Factor 2", 
     ylim = c(-1,1),
     xlim = c(-1,1),
     main = "Varimax rotation")
     abline(h = 0, v = 0)
#Note: If you also want to put the variable label here, just take out the #.
#load <- df_factor_var$loadings[,1:2]
#text(text(load,labels=names(df),cex=.7, col="red"))


# Oblimin Rotation
plot(df_factor_obl$loadings[,1], 
     df_factor_obl$loadings[,2],
     xlab = "Factor 1", 
     ylab = "Factor 2",
     ylim = c(-1,1),
     xlim = c(-1,1),
     main = "Oblimin rotation")
abline(h = 0, v = 0)
#load <- df_factor_obl$loadings[,1:2]
#text(text(load,labels=names(df),cex=.7, col="yellow"))
#abline(h = 0, v = 0)


knitr::spin(hair = "Factor_Analysis_TDMLecture.R")

