---
title: "Cluster Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "README_files/3-ClusterAnalysis/")
```

## Example Airports

### Variables: 

`Code`
`Airport`
`Ordem`
`Passengers`
`Movements`
`Number of airlines`
`Main air line flights percentage`
`Maximum percentage of traffic per country`
# Number of Low Cost Airlines
# Low Cost Airlines percentage
# Destinations
# Average_route_Distance
# Distance to closest Similar Airport
# Airport Regional Relevance
# Distance to city km
# Inhabitants corrected
# Number of visitors corrected
# GDP corrected
# Cargoton


#### Import Libraries

```{r}
library(readxl) # Library used for reading excel files
library(skimr) # Library used for summary statistics
library(tidyverse) # Library used in data science to perform exploratory data analysis
library(mclust) # Library used for model based clustering
library(cluster) # Library used for cluster analysis
library(factoextra) # Library used for visualizing distances
```

#### Set working directory

```{r}
setwd("G:/O meu disco/TDM - Lecture R/TDM github/Transport-Demand-Modelling/")
```

#### Import dataset

```{r}
dataset <- read_excel("Data/Data_Aeroports_Clustersv1.xlsX")
```

#### Transform dataset into dataframe

```{r}
df <- data.frame(dataset)
```

#### Summary statistics

```{r}
skim(df)
```

#### Now let us plot an example and take a look

```{r}
plot(Numberofairlines ~ Destinations, df)
with(df, text(Numberofairlines ~ Destinations, label = Airport, pos = 4, cex = 0.6))
```

> **Note:** You can already guess the number of clusters by visualizing the two variables. However, this is not clear and it does not consider the other variables. 

#### Treat the data before performing a cluster analysis

* In this example we do not have missing values. In case you do have in the future, you can take out the missing values with listwise deletion.

```{r}
 df <- na.omit(df)
```

* Leave only continuous variables, and take out "Ordem"

```{r}
  drop <- c("Code","Airport", "Ordem")
  df_reduced = df[,!(names(df) %in% drop)]
```

Analyze how the scale of the values of the variables are different 
  
```{r}
head(df_reduced)
```

* Standardization (Z-score) - (xi - xmean / standard deviation)

```{r}
  mean <- apply(df_reduced, 2, mean) # The "2" in the function is used to select the columns. MARGIN: c(1,2)
  sd <- apply(df_reduced, 2, sd)
  df_scaled <- scale(df_reduced, mean, sd)
```
  
### HIERARCHICAL CLUSTERING

#### Measuring Similarity through Euclidean distances

```{r}
  distance <- dist(df_scaled, method = "euclidean")
  print(distance, digits = 3)
```
  
#### Visualize distances in heatmap

```{r}
  fviz_dist(distance, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"), order = FALSE)
```

> **Note:** There are other forms distance measures that can be used such as: 
  i) Minkowski distance; 
  ii) Manhattan distance; 
  iii) Mahanalobis distance.


**1. Complete linkage (Farthest neighbor) clustering algorithm**

Based on the maximum distance between observations in each cluster.

```{r}
modelc <- hclust(distance, "complete")
plot(modelc, labels = df$Airport, xlab = "Distance - Complete linkage", hang = -1)

# Visualize the cut on the tree 
rect.hclust(modelc, 4, border = "blue") 
```

**2. Average linkage between groups** 

The distance between clusters is the average of the distances between observations in one cluster to all the members in the other cluster. 


```{r}
modela <- hclust(distance, "average")
plot(modela, labels = df$Airport, xlab = "Distance - Average linkage", hang = -1)
rect.hclust(modelc, 4, border = "red")
```

**3. Ward`s method**
  
The measures of similarity are the sum of squares within the cluster summed over all variables. 


```{r}
modelw <- hclust(distance, "ward.D2")
plot(modelw, labels = df$Airport, xlab = "Distance - Ward method", hang = -1)
# Visualize where to cut on the tree (choose number of clusters)
rect.hclust(modelw, 4, border = "orange")
```

**4. Centroid method**

The similarity between two clusters is the distance between its centroids.
  
```{r}
modelcen <- hclust(distance, "centroid")
plot(modelcen, labels = df$Airport, xlab = "Distance - Centroid method", hang = -1)
rect.hclust(modelcen, 4, border = "green")
```

#### Now lets evaluate the membership of each observation with the cutree function for each method.

```{r}
member_com <- cutree(modelc, 4)
member_av <- cutree(modela, 4)
member_ward <- cutree(modelw, 4)
member_cen <- cutree(modelcen, 4)
```
  

#### Plot table to compare how common each method is to each other.

Let us compare the complete linkage with the average linkage

```{r}
table(member_com, member_av)
```
  
  
#### Silhouette Plot 

  plot(silhouette(member_com, distance))
  plot(silhouette(member_av, distance))
  plot(silhouette(member_ward, distance))
  plot(silhouette(member_cen, distance))

>**Note:** Analyzes how similiar an observation is to its own cluster compared to other clusters. The clustering configuration is appropriate when most objects have high values. Low or negative values indicate that the clustering does not have an appropriate number of clusters.

### NON-HiERARCHICAL CLUSTERING 

#### K-means clustering

```{r}
km_clust <- kmeans(df_scaled, 3)
```

Print out the results

```{r}
km_clust 

str(km_clust)

```


#### Choosing K 

#### This algorithm will detect how many clusters from 1 to 10 explains more variance
  
```{r}
  k <- list()
  for(i in 1:10){
    k[[i]] <- kmeans(df_scaled, i)
  }
```
  
#### Print the k value and take a look at the ratio (between_SS / total_SS)

```{r}
k
```

#### Now lets try to plot (between_SS / total_SS) in to a scree plot

```{r}
betSS_totSS <- list()
for(i in 1:10){
betSS_totSS[[i]] <- k[[i]]$betweenss/k[[i]]$totss
}
plot(1:10, betSS_totSS, type = "b", ylab = "Between SS / Total SS", xlab = "Number of clusters")
```


#### Finally, try plotting each variable with each other and analyze if the clusters make sense.  

Let us go back to first example and take a look. 

```{r}
plot(Numberofairlines ~ Destinations, df, col = km_clust$cluster)
with(df, text(Numberofairlines ~ Destinations, label = Airport, pos = 1, cex = 0.6))
```



