---
title: "Generalized Linear Models"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "README_files/GLM/")
```

## Example: Accident data from California (1993 to 1998) and Michigan (1993 to 1997) 

> Your Task: Analyze the number o accidents.

### Variables:

`ACCIDENT`: Count on injury accidents over observation period;  

`STATE`: indicator variable for state:  

    0: California  
    
    1: Michigan  
    
`AADT1`: Average annual daily traffic on major road;  

`AADT2`: Average annual daily traffic on minor road;  

`Median`: Median width on major road in feet;  

`DRIVE`: Number of driveways within 250 ft of intersection center.

#### Import Libraries

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(readxl) # Library used for reading excel files
library(skimr) # Library used for summary statistics
library(tidyverse) # Library used in data science to perform exploratory data analysis
library(MASS) # Library used for negative binomial regression
library(vcd) # Library used for goodness of fit parameters
library(car) # Library used for goodness of fit
library(rcompanion) # Library used for goodness of fit
library(popbio) # Library used to calculate elasticities
```

#### Set working directory

```{r}
setwd("G:/O meu disco/TDM - Lecture R/TDM github/Transport-Demand-Modelling")
```

#### Import dataset

```{r}
dataset <- read_excel("Data/TDM_GZLM_CALMICH_Example.xlsX")
```

#### Let us take a look at the summary and structure of the dataset

```{r}
summary(dataset)
str(dataset)
```

> **Note**: The dataset looks weird. Use the view function to take a better look at the dataset.

```{r echo=TRUE}
view(dataset)
```

#### Treat the dataset

#### The best way to treat the dataset is to import it again. Use the following steps:  

   **1.** Go to "Import Dataset" on the "Environment" window at the upper right display;  
   
   **2.** Click on "From excel";  
   
   **3.** Check if "First row as names" is checked;  
   
   **4.** Put the number of rows you want to skip and click on "import";  
   
   **5.** This will generate a code, which you can copy and use it the next time you open the file.  
   
#### Therefore, here is the following code:

```{r}
TDM_GZLM_CALMICH_Example <- read_excel("Data/TDM_GZLM_CALMICH_Example.xlsx", skip = 5)
```
 
#### Transform "TDM_GZLM_CALMICH_Example" into a dataframe to better manage the dataset

```{r}
df <- data.frame(TDM_GZLM_CALMICH_Example)
view(df)
```
 
#### Now we can analyze the descriptive statistics of the dataframe
 
```{r}
str(df)
skim(df)
```
 
#### Take a look at the histograms of the variables

```{r}
hist(df$ACCIDENT, xlab = "ACCIDENT", main = "Histogram of ACCIDENT") 
hist(df$STATE, xlab = "STATE", main = "Histogram of STATE")
hist(df$AADT1, xlab = "AADT1", main = "Histogram of AADT1")
hist(df$AADT2, xlab = "AADT2", main = "Histogram of AADT2")
hist(df$MEDIAN, xlab = "MEDIAN", main = "Histogram of MEDIAN")
hist(df$DRIVE, xlab = "DRIVE", main = "Histogram of DRIVE")
```

#### Plot the density function of the variable ACCIDENTS

```{r}
plot(density(df$ACCIDENT), main="Density estimate of ACCIDENTS")
```

> **Note**: As the dependent variable is "count data", as has discrete values, then a poisson distribution should be more adequate. 

#### Poisson assumption: 
Take a look at the mean and the variance of the Dependent variable. Check if they are equal to each other. 

```{r}
mean(df$ACCIDENT)
var(df$ACCIDENT)
```

Or you could calculate the coefficient of variance:

```{r}
var(df$ACCIDENT)/mean(df$ACCIDENT)
```

> **Note**: If the coefficient of variance > 1, then you have overdispersion.

#### Try estimating goodness of fit parameter for the PDF of ACCIDENT. Use the Maximum Likelihood method.

```{r}
gf<-goodfit(df$ACCIDENT,type= "poisson", method= "ML")
summary(gf)
```

> **Note**: The null hypothesis is that it is a Poisson distribution. Therefore, for it to be a poisson distribution, the pvalue > 0.05.

#### Now let us run the many possible models  

**1. Poisson model:**
 
```{r}
model1 = glm(ACCIDENT ~ as.factor(STATE) + AADT2 + MEDIAN + DRIVE + offset(log(AADT1)), family = poisson(link = "log"), data = df, method = "glm.fit")
```
 
> **Note**: The method "glm.fit" uses iteratively reweighted least squares to fit the model. Try looking for other methods and see the difference. 
  
There are many families and links that can be used, depending on the characteristics of your data.

```{r echo=FALSE}
family <-c ("Gaussian", "Binomial", "Poisson", "Gamma", "inverse.gaussian")
link <-c("identity", "logit, probit, cloglog", "log, identity, sqrt", "inverse, identity, log", "1/mu^2")
table <- data.frame(family,link)
knitr::kable(table, align = "l")
```

```{r}
summary(model1)
```

If we obtain:  

residuals > degrees of freedom (overdispersion);  

residuals < degrees of freedom (underdispersion);  

residuals = degrees of freedom (mean = variance).  

> **Note:** In overdispersion, the estimates are reliable but the standard errors tend to be smaller. 
 
#### Calculate the pseudo-Rsquare and perform an Omnibus test

```{r}
nagelkerke(model1)
```

> **Note**: The likelihood ratio test (Omnibus test) compares the fitted model ("Model") with the only-intercept model ("Null"). This test verifies if the explained variance is higher than the the unexplained variance.

> **Note**: Ho: There is no overdispersion in the model. Therefore, if pvalue < 0.05, there is overdispersion, and we should choose to use a Negative Binomial model.

#### Calculate the Type III test.

```{r}
Anova(model1, type = "III", test = "Wald")
```

> **Note**: Type III tests examine the significance of each partial effect. Thus, it considers the significance of an effect with all the other effects in the model. The Chisq tests the significance of the effect added to the model by having all of the other effects.

#### Let us correct the standard errors with an overdispersed poisson 
 
```{r}

model2 = glm(ACCIDENT ~ as.factor(STATE) + AADT2 + MEDIAN + DRIVE + offset(log(AADT1)), family = quasipoisson(link = "log"), data = df, method = "glm.fit")

summary(model2)
```

> **Note**: The estimates are the same, but the standard errors have increased because they are adjusted by the scale parameter

#### Let us try the negative binomial distribution
 
```{r}
model3 = glm.nb(ACCIDENT ~ as.factor(STATE) + AADT2 + MEDIAN + DRIVE + offset(log(AADT1)), data = df)

summary(model3)
```

#### Calculate the pseudo-Rsquare and perform an Omnibus test   

```{r}
Anova(model3, type = "III", test = "Wald") 
```

#### Calculate the Type III test.  

```{r}
nagelkerke(model3)
```

#### Compare model 1 and model 3:
 
#### Calculate the Akaike’s Information Criteria (AIC) and the Bayesian Information Criteria (BIC) 


```{r}
aic <- data.frame(model1 = AIC(model1), model3 = AIC(model3))
knitr::kable(aic, align = "l")
```

```{r}
bic <- data.frame(model1 = BIC(model1), model3 = BIC(model3))
knitr::kable(bic, align = "l")
```

> **Note**: AIC and BIC evaluates the quality of a finite set of models.
  
> **Note**: AIC and BIC consider the maximum likelihood and the number of parameters in assessing the quality of the models.Nonetheless, the difference between both methods is that the BIC takes into account the number of observations of dataset. 
  
> **Note**: The smaller the values of AIC and BIC, the better the model 

#### Calculate the elasticities of the negative binomial model (model 3)

```{r}
el1 <- as.numeric(model3$coefficients["AADT1"] * mean(df$AADT1)/mean(df$ACCIDENT))
el2 <- as.numeric(model3$coefficients["AADT2"] * mean(df$AADT2)/mean(df$ACCIDENT))
el3 <- as.numeric(model3$coefficients["MEDIAN"] * mean(df$MEDIAN)/mean(df$ACCIDENT))
el4 <- as.numeric(model3$coefficients["DRIVE"] * mean(df$DRIVE)/mean(df$ACCIDENT))
el5 <- as.numeric(model3$coefficients["STATE"] * mean(df$STATE)/mean(df$ACCIDENT))

variable <-c ("AADT1", "AADT2", "MEDIAN", "DRIVE", "STATE")
elasticity <-c (el1, el2, el3, el4, el5)
elas_table <- data.frame(variable,elasticity)
knitr::kable(elas_table, align = "l")
```
  
  
> **Note:** AADT1 does not have a value because it is the offset of the model. Additionally, the variable STATE also does not have a value because it is a categorical variable. In this case, it would be necessary to calculate the pseudo-elasticity. 

