---
title: "Hazard-Based Duration Models"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "README_files/8-HazardBasedModels/")
```

## Example: Work-to-home departure delay

A survey of 204 Seattle-area commuters was conducted to examine the duration of time that commuters delay their work-to-home trips in an effort to avoid peak period traffic congestion. Of the 204 commuters surveyed, 96 indicated that they sometimes delayed their work-to-home trip to avoid traffic congestion. These commuters provided their average time delay. Thus, each commuter has a completed delay duration so that neither left nor right censoring is present in the data.

> **Your task:** i) Plot the Kaplan-Meier estimate of the duration of time that commuters delay
their work-to-home trips; ii) Determine the significant factors that affect the duration of commuters’ delay
using a Cox model; iii) Examine the work-to-home departure delay using exponential, Weibull, and
log-logistic proportional-hazards models.

#### Import Libraries

```{r}
library(readxl)
library(skimr)
library(survival)
library(coxme)
library(survminer)
```

#### Set working directory

```{r}
setwd("G:/O meu disco/TDM - Lecture R/TDM github/Transport-Demand-Modelling")
```

# Import dataset, tranform in dataframe, and take a first look.

```{r}
data.delay <- read_excel("Data/ExerciseHBDM.xlsx")

data.delay <- data.frame(data.delay)

head(data.delay)
```

>**Note:** The variables do not have names. Use the algorithm below to create new columns with the assigned variables.

#### Create new variables and assign the values to each one respectively. 

```{r}
data.delay["minutes"] <- NA

data.delay$minutes <-  data.delay$X1

data.delay["activity"] <- NA

data.delay$activity <-  data.delay$X2

data.delay["number_of_times"] <- NA

data.delay$number_of_times <-  data.delay$X3

data.delay["mode"] <- NA

data.delay$mode <-  data.delay$X4

data.delay["route"] <- NA

data.delay$route <-  data.delay$X5

data.delay["congested"] <- NA

data.delay$congested <-  data.delay$X6

data.delay["age"] <- NA

data.delay$age <-  data.delay$X7

data.delay["gender"] <- NA

data.delay$gender <-  data.delay$X8

data.delay["number_cars"] <- NA

data.delay$number_cars <-  data.delay$X9

data.delay["number_children"] <- NA

data.delay$number_children <-  data.delay$X10

data.delay["income"] <- NA

data.delay$income <-  data.delay$X11

data.delay["flexible"] <- NA

data.delay$flexible <-  data.delay$X12

data.delay["distance"] <- NA

data.delay$distance <-  data.delay$X13

data.delay["LOSD"] <- NA

data.delay$LOSD <-  data.delay$X14

data.delay["rate_of_travel"] <- NA

data.delay$rate_of_travel <-  data.delay$X15

data.delay["population"] <- NA

data.delay$population <-  data.delay$X16

data.delay["retail"] <- NA

data.delay$retail <-  data.delay$X17

data.delay["service"] <- NA

data.delay$service <-  data.delay$X18

data.delay["size"] <- NA

data.delay$size <-  data.delay$X19    
```

#### Take a look at the structure
```{r}

str(data.delay)
```

>**Note:** We have repeated variables. Let us exclude the original ones without labels. 

#### Exclude original variables

```{r}
drop <- c("X1","X2", "X3", "X4", "X5", "X6", "X7", "X8", "X9", "X10", "X11", "X12", "X13", "X14", "X15", "X16", "X17", "X18", "X19")
df = data.delay[,!(names(data.delay) %in% drop)]
```

```{r}
skim(df)
```


#### Sort the data by time
```{r}
df <- df[order(df$minutes),]
print(df)

```

#### Create graph
```{r}
with(df, plot(minutes, type="h"))
```

#### Create the life table survival object for df

```{r}
data.delay2 <-subset(df, minutes>0)
df.survfit = survfit(Surv(minutes) ~ 1, data= data.delay2)
summary(df.survfit)
```

>**Note:** The functions survfit() and Surv() create a life table survival object.

#### Plot the Kaplan-Meier curve

```{r}
plot(df.survfit, xlab = "Time (minutes)", ylab="Survival
probability", conf.int=TRUE)
ggsurvplot(df.survfit, xlab = "Time (minutes)", xlim =
range(0:250) , conf.int = TRUE, color = "red", ggtheme =
theme_minimal())
```

#### Cox Proportional Hazard Model Estimates of the Duration of Commuter Work-To-Home Delay to Avoid Congestion

```{r}
result.cox <- coxph(Surv(minutes) ~ gender + rate_of_travel + distance + population, data= data.delay2)
summary(result.cox)
```

#### Testing proportional Hazards assumption

```{r}
test.ph <- cox.zph(result.cox)
plot(test.ph)
ggcoxzph(test.ph)
```

> **Note:** Include an interaction between the covariate and a function of time (or distance). Log time often used but could be any function. If significant then assumption violated.

> **Note:** Test the proportional hazards assumption on the basis of partial residuals. Type of residual known as Schoenfeld residuals.

> **Note:** For each covariate, the function cox.zph() correlates the corresponding set of scaled Schoenfeld residuals with time, to test for independence between residuals and time. Additionally, it performs a global test for the model as a whole.

> **Note:** In principle, the Schoenfeld residuals are independent of time. A plot that shows a non-random pattern against time is evidence of violation of the PH assumption.

#### Plot the baseline survival function

```{r}
#  slide 56
```


