---
title: "Hazard-Based Duration Models"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "README_files/8-HazardBasedModels/")
```

### Example: Work-to-home departure delay

A survey of 204 Seattle-area commuters was conducted to examine the duration of time that commuters delay their work-to-home trips in an effort to avoid peak period traffic congestion. Of the 204 commuters surveyed, 96 indicated that they sometimes delayed their work-to-home trip to avoid traffic congestion. These commuters provided their average time delay. Thus, each commuter has a completed delay duration so that neither left nor right censoring is present in the data.

**Your task:**

1. Plot the Kaplan-Meier estimate of the duration of time that commuters delay
their work-to-home trips;
2. Determine the significant factors that affect the duration of commuters’ delay
using a Cox model;
3. Examine the work-to-home departure delay using exponential, Weibull, and
log-logistic proportional-hazards models.

## Get to know your data
##### Import Libraries

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(readxl)
library(skimr)
library(survival)
library(coxme)
library(survminer)
library(ggplot2)
```

##### Import dataset, transform in dataframe, and take a first look.

```{r}
data.delay <- read_excel("Data/ExerciseHBDM.xlsx")
data.delay <- data.frame(data.delay)
head(data.delay)
```

##### Assign variable lables to each one respectively
```{r}
names(data.delay) <- c("id","minutes","activity","number_of_times","mode","route","congested","age",
"gender","number_cars","number_children","income","flexible","distance",
"LOSD","rate_of_travel","population","retail","service","size") 
data.delay <- data.frame(data.delay, row.names = 1) #make id (1st variable) as row name
```

##### Take a look at the structure
```{r}
str(data.delay)
```

```{r}
df <- data.delay
skim(df)
```


##### Sort the data by time
```{r}
df <- df[order(df$minutes),]
```

##### Plot yout data
```{r}
plot(df$minutes, type="h")
```

## Survival function
##### Create the life table survival object for df

```{r}
data.delay2 <-subset(df, minutes>0)
df.survfit = survfit(Surv(minutes) ~ 1, data= data.delay2)
summary(df.survfit)
```

>**Note:** The functions `survfit()` and `Surv()` create a life table survival object.

##### Plot the Kaplan-Meier curve

```{r}
# option 1
plot(
  df.survfit,
  xlab = "Time (minutes)",
  ylab = "Survival
probability",
conf.int = TRUE
)
```
```{r}
#option 2
ggsurvplot(
  df.survfit,
  xlab = "Time (minutes)",
  xlim =
    range(0:250) ,
  conf.int = TRUE,
  pallete = "red",
  ggtheme =
    theme_minimal()
)
```


> **Note:** It is the most widely applied nonparametric method in survival analysis.The Kaplan–Meier method provides useful estimates of survival probabilities and a graphical presentation of the survival distribution.  
The Kaplan–Meier method assumes that censoring is independent of survival times. If this is false, the Kaplan–Meier method is inappropriate.

#### Cox Proportional Hazard Model Estimates of the Duration of Commuter Work-To-Home Delay to Avoid Congestion

```{r}
result.cox <- coxph(Surv(minutes) ~ gender + rate_of_travel + distance + population, data= data.delay2)
summary(result.cox)
```

> **Note:**
The Cox proportional-hazards model is semiparametric method that produces estimated hazard ratios (sometimes called rate ratios or risk ratios).  
Regression coefficients are on a log scale.

##### Testing proportional Hazards assumption

```{r}
test.ph <- cox.zph(result.cox)
plot(test.ph)
ggcoxzph(test.ph)
```

> **Note:**
It includes an interaction between the covariate and a function of time (or distance). Log time is often used but it could be any function. If significant then the assumption is violated.  
Test the proportional hazards assumption on the basis of partial residuals. Type of residual known as Schoenfeld residuals.  
For each covariate, the function cox.zph() correlates the corresponding set of scaled Schoenfeld residuals with time, to test for independence between residuals and time. Additionally, it performs a global test for the model as a whole.  
In principle, the Schoenfeld residuals are independent of time. A plot that shows a non-random pattern against time is evidence of violation of the PH assumption.

##### Plot the baseline survival function

```{r}
ggsurvplot(
  survfit(result.cox),
  data = data.delay2,
  palette = "#2E9FDF",
  ggtheme = theme_minimal()
)
```

##### Plot the cummulative hazard function

```{r}
ggsurvplot(
  survfit(result.cox),
  data = data.delay2,
  conf.int = TRUE,
  palette = c("#FF9E29", "#86AA00"),
  risk.table = TRUE,
  risk.table.col = "strata",
  fun = "event"
)
```

##### Calculate the Log-likelihood




