---
eval: true
code-fold: false
number-depth: 2
engine: knitr
---

# R basics

In this chapter we will introduce to the R basics and some exercises to get familiar to how R works.

## Math operations

### Sum

```{r}
1+1
```

### Subtraction

```{r}
5-2
```

### Multiplication

```{r}
2*2
```

### Division

```{r}
8/2
```

### Round the number

```{r}
round(3.14)
round(3.14, 1) # The "1" indicates to round it up to 1 decimal digit.
```

You can use help `?round` in the console to see the description of the function, and the default arguments.

## Basic shortpaths

### Perform Combinations

```{r}
c(1, 2, 3)
c(1:3) # The ":" indicates a range between the first and second numbers. 
```

### Create a comment with `ctrl + shift + m`

```{r}
# Comments help you organize your code. The software will not run the comment. 
```

### Create a table

A simple table with the number of trips by car, PT, walking, and cycling in a hypothetical street segment at a certain period.

**Define variables**

```{r}
modes <- c("car", "PT", "walking", "cycling") # you can use "=" or "<-"
Trips = c(200, 50, 300, 150) # uppercase letters modify

```

**Join the variables to create a table**

```{r}
table_example = data.frame(modes, Trips)
```

**Take a look at the table**

Visualize the table by clicking on the "Data" in the "Environment" page or use :

```{r}
View(table_example)
```

**Look at the first row**

```{r}
table_example[1,] #rows and columns start from 1 in R, differently from Python which starts from 0.
```

**Look at first row and column**

```{r}
table_example[1,1]
```

## Practical exercise

**Dataset:** the number of trips between all municipalities in the Lisbon Metropolitan Area, Portugal [@IMOB].

### Import dataset

You can click directly in the file under the "Files" pan, or:

```{r}
data = readRDS("data/TRIPSmode.Rds")
```

::: {.callout-tip appearance="simple"}
After you type `"` you can use `tab` to navigate between folders and files and `enter` to autocomplete.
:::

### Take a first look at the data

**Summary statistics**

```{r}
summary(data)
```

**Check the structure of the data**

```{r}
str(data)
```

**Check the first values of each variable**

```{r}
#| eval: false
data
```

```{r}
head(data, 3) # first 3 values
```

**Check the number of rows (observations) and columns (variables)**

```{r}
nrow(data)
ncol(data)
```

**Open the dataset**

```{r}
View(data)
```

### Explore the data

**Check the total number of trips**

Use `$` to select a variable of the data

```{r}
sum(data$Total)
```

**Percentage of car trips related to the total**

```{r}
sum(data$Car)/sum(data$Total) * 100
```

**Percentage of active trips related to the total**

```{r}
(sum(data$Walk) + sum(data$Bike)) / sum(data$Total) * 100
```

### Modify original data

**Create a column with the sum of the number of trips for active modes**

```{r}
data$Active = data$Walk + data$Bike
```

**Filter by condition (create new tables)**

Filter trips only with origin from Lisbon

```{r}
data_Lisbon = data[data$Origin == "Lisboa",]
```

Filter trips with origin **different** from Lisbon

```{r}
data_out_Lisbon = data[data$Origin != "Lisboa",]
```

Filter trips with origin **and** destination in Lisbon

```{r}
data_in_Out_Lisbon = data[data$Origin == "Lisboa" & data$Destination == "Lisboa",]
```

**Remove the first column**

```{r}
data = data[ ,-1] #first column
```

**Create a table only with origin, destination and walking trips**

There are many ways to do the same operation.

```{r}
names(data)
```

```{r}
data_walk2 = data[ ,c(1,2,4)]
```

```{r}
data_walk3 = data[ ,-c(3,5:9)]
```

### Export data

Save data in **.csv** and **.Rds**

```{r}
write.csv(data, 'data/dataset.csv', row.names = FALSE)
saveRDS(data, 'data/dataset.Rds') #Choose a different file. 
```

### Import data

```{r}
csv_file = read.csv("data/dataset.csv")
rds_file = readRDS("data/dataset.Rds")
```

```{r}
#| include: false
#| eval: false
# this coverts this quarto to a plain r script
knitr::purl("r-basics.qmd", "code/r-basics.R", documentation = 2)

```
