---
title: "Multiple Linear Regression"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**INSTITUTO SUPERIOR TECNICO, UNIVERSITY OF LISBON**  
**TRANSPORT DEMAND MODELLING COURSE, PROF. FILIPE MOURA**

## Chicago example exercise

Trip production of 57 Traffic Assignment Zones of Chicago in 1960's

> Your task: Estimate a linear regression model that predicts trips per occupied dwelling unit.  

### Variables:

* `TODU`: Motorized Trips (private car or Public Transportation) per occupied dwelling unit;
* ACO: Average car ownership (cars per dwelling);
* AHS: Average household size;
* SRI: Social Rank Index:  
      1. proportion of blue-collar workers (e.g., construction, mining);  
      2. proportion of people with age higher than 25 years that have completed at least 8 year of education;  
**Note:** The SRI has its maximum value when there are no blue-collar workers and all adults have education of at least 8 years. 

> UI: Urbanization Index:
1. fertility rate, defined as the ratio of children under 5 years of age to  the female population of childbearing age;  
2. female labor force participation rate, meaning the % of women who are in the labor force;  
3. % of single family units to total dwelling units.

      ## The degree of urbanization index would be increased by
        ### a) lower fertility rate,
        ### b) higher female labor force participation rate, and
        ### c) higher proportion of single dwelling units.

      ## Note: High values for this index imply less attachment to the home

SI:Segregation Index
      ## It measures the proportion of an area to which minority groups
      ## (e.g: non-whites, foreign-born, Eastern Europeans) live in isolation.
      
      ## Note: High values for this index imply that those communities are less
      ## prone to leaving their living areas and as such to having lower
      ## levels of mobility.



### Import Libraries
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(skimr)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
