---
title: "RMarkdown reports"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
_work in progress_

How to create static or dynamic reports for your home assignments, using R markdown.

R Markdown provides an authoring framework for data science. You can use a single R Markdown file to both.

Generate high quality reports that can be shared with an audience.
R Markdown documents are fully reproducible and support dozens of static and dynamic output formats.


# R markdown

First install pandoc (https://pandoc.org/installing.html)

The library `rmarkdown` already comes with RStudio.
If you neet to install it:
```r
install.packages("rmarkdown")
```

## Usefull resources
* [**A great quick tutorial**](https://towardsdatascience.com/getting-started-in-r-markdown-2d3de636bde3)
* [R Markdown Cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/rmarkdown-2.0.pdf)
* [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)


### Make your outputs live
To make your outputs online at your IST server.

#### VPN check

First, you have to make sure that your computer is trusted to be connected to IST servers.  
Please [follow the insctructions](https://si.tecnico.ulisboa.pt/en/servicos/redes-e-conetividade/vpn/), by choosing your Operating System.

You only need to do this once per computer.


#### Using SFTP

SFTP is a Secure FTP connection.  
Donwload FileZilla at: https://filezilla-project.org/download.php?type=client

* CTRL + S (site manager)
* new site (name "fenix")
* SFTP
* server: sigma.ist.utl.pt
* port: _leave blank_
* user: ist1XXXXX
* Connect, save password, always trust this key

Upload your outputs under **web** / tdm



```{r knit, eval=FALSE, include=FALSE}
#para colocar os ficheiros nos locais certos (> ver batch process em knit.R)
name = "00-RMarkdownReports"
file.rename(paste0("RmdFiles/", name, ".md"), paste0(name, ".md"))
```