---
title: "Hazard-Based Duration Models"
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE, purl=F}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "9-HazardBasedModels/")
```

#### Example: Work-to-home departure delay

A survey of 204 Seattle-area commuters was conducted to examine the duration of time that commuters delay their work-to-home trips in an effort to avoid peak period traffic congestion. Of the 204 commuters surveyed, 96 indicated that they sometimes delayed their work-to-home trip to avoid traffic congestion. These commuters provided their average time delay. Thus, each commuter has a completed delay duration so that neither left nor right censoring is present in the data.

**Your tasks:**

1. Plot the Kaplan-Meier estimate of the duration of time that commuters delay
their work-to-home trips;
2. Determine the significant factors that affect the duration of commuters’ delay
using a Cox model;
3. Examine the work-to-home departure delay using exponential, Weibull, and
log-logistic proportional-hazards models.

## Get to know your data
##### Import Libraries

```{r message=FALSE, warning=FALSE}
library(readxl)
library(skimr)
library(survival)
library(coxme)
library(survminer)
library(ggplot2)
```

##### Import dataset, transform in dataframe, and take a first look.

```{r}
data.delay <- read_excel("Data/ExerciseHBDM.xlsx")
data.delay <- data.frame(data.delay)
head(data.delay) #variable names are missing
```

##### Assign variable lables to each one respectively
```{r}
names(data.delay) <- c("id","minutes","activity","number_of_times","mode","route","congested","age",
                        "gender","number_cars","number_children","income","flexible","distance",
                        "LOSD","rate_of_travel","population","retail","service","size") 
df <- data.frame(data.delay, row.names = 1) #make id (1st variable) as row name
```

##### Take a look at the structure
```{r}
str(df)
skim(df)
```

##### Plot yout data
```{r}
df <- df[order(df$minutes),] #sort by time
plot(df$minutes, type="h") #high-density vertical lines
```

## Survival function
### 1.Plot the Kaplan-Meier estimate curve
##### Create the life table survival object for df

```{r}
data.delay2 <-subset(df, minutes>0)
df.survfit = survfit(Surv(minutes) ~ 1, data= data.delay2)
summary(df.survfit)
```

>**Note:** The functions `survfit()` and `Surv()` create a life table survival object.

##### Plot the Kaplan-Meier curve

```{r}
# option 1
plot(
  df.survfit,
  xlab = "Time (minutes)",
  ylab = "Survival
probability",
conf.int = TRUE
)
```
```{r}
#option 2
ggsurvplot(
  df.survfit,
  xlab = "Time (minutes)",
  xlim =
    range(0:250) ,
  conf.int = TRUE,
  pallete = "red",
  ggtheme =
    theme_minimal()
)
```


> **Note:** It is the most widely applied nonparametric method in survival analysis.The Kaplan–Meier method provides useful estimates of survival probabilities and a graphical presentation of the survival distribution.  
The Kaplan–Meier method assumes that censoring is independent of survival times. If this is false, the Kaplan–Meier method is inappropriate.

### 2. Cox proportional-hazards Model  
The Cox proportional-hazards model is semiparametric method that produces estimated hazard ratios (sometimes called rate ratios or risk ratios).  
Let's perform a Cox Proportional Hazard Model Estimate estimate the model for Duration of Commuter Work-To-Home Delay to Avoid Congestion

```{r}
result.cox <- coxph(Surv(minutes) ~ gender + rate_of_travel + distance + population, data= data.delay2)
summary(result.cox)
```

> **Note:** Regression coefficients are on a log scale.

##### Testing proportional Hazards assumption  
Test the proportional hazards assumption on the basis of partial residuals. Type of residual known as Schoenfeld residuals.  
It includes an interaction between the covariate and a function of time (or distance). Log time is often used but it could be any function.
For each covariate, the function `cox.zph()` correlates the corresponding set of scaled Schoenfeld residuals with time, to test for independence between residuals and time. Additionally, it performs a global test for the model as a whole.  
A plot that shows a non-random pattern against time is evidence of violation of the PH assumption.  
```{r}
test.ph <- cox.zph(result.cox)
par(mfrow=c(2,2))
plot(test.ph)
ggcoxzph(test.ph)
```

> **Note:**  If significant then the assumption is violated.  
In principle, the Schoenfeld residuals are independent of time.


##### Plot the baseline survival function

```{r}
ggsurvplot(
  survfit(result.cox),
  data = data.delay2,
  palette = "#2E9FDF",
  ggtheme = theme_minimal()
)
```

##### Plot the cummulative hazard function

```{r}
ggsurvplot(
  survfit(result.cox),
  data = data.delay2,
  conf.int = TRUE,
  palette = c("#FF9E29", "#86AA00"),
  risk.table = TRUE,
  risk.table.col = "strata",
  fun = "event"
)
```

##### Calculate the McFadden Pseudo R-square

```{r}
result.cox$loglik
LLi = result.cox$loglik[1] # Initial log-likelihood
LLf = result.cox$loglik[2] # Final log-likelihood
1- (LLf/LLi) # pseudo R-suare
```


### 3. Exponential, Weibull, and log-logistic proportional-hazards models
Parametric Model Estimates of the Duration of Commuter Work-ToHome Delay to Avoid Congestion

* **Exponential**
```{r}
survreg(
  Surv(minutes) ~ gender + rate_of_travel + distance + population,
  data = data.delay2,
  dist = "exponential" )
```

* **Weibull**
```{r}
survreg(
  Surv(minutes) ~ gender + rate_of_travel + distance + population,
  data = data.delay2,
  dist = "weibull" )
```

* **Log-logistic**
```{r}
survreg(
  Surv(minutes) ~ gender + rate_of_travel + distance + population,
  data = data.delay2,
  dist = "loglogistic" ) 
```

>**Note:** The argument _dist_ has several options to describe the parametric model used ("weibull", "exponential", "gaussian", "logistic", "lognormal", or "loglogistic". See more with `?survreg`
